---
title: \sf Updated_GATA3_ChIP_motif_analysis
header-includes:
- \usepackage{color}
- \usepackage{float}
- \DeclareUnicodeCharacter{2212}{-}
date: "March 04, 2024"
output:
  bookdown::html_document2:
    toc: true
fontsize: 14pt
geometry: margin=1in
---

```{css, echo=FALSE}
body .main-container {
  max-width: 1200px !important;
  width: 1200px !important;
}
body {
  max-width: 1200px !important;
}

pre {
  max-height: 600px;
  overflow-y: auto;
}

pre[class] {
  max-height: 600px;
}
```

```{css, echo=FALSE}
.watch-out {
  background-color: lightcyan;
  border: 3px solid lightgrey;
  font-weight: bold;
}

.revision {
  background-color: #FFF4F2;
  border: 3px solid lightgrey;
  font-weight: bold;
}

.highlighted
{
   color:red;
}

.highlighted2
{
   color:blue;
}
```

```{r engine='R', eval=TRUE, echo=F}
knitr::opts_chunk$set(class.source = "watch-out")
```


# Install required packages
Install `bigWig` and `latticeExtra` package \
```{r engine='R', eval=F, echo=T}
install.packages("devtools", quiet = TRUE)
library(devtools)
devtools::install_github('andrelmartins/bigWig',
              subdir='bigWig')
library(bigWig)

install.packages("DESeq2", quiet = TRUE)
install.packages("dplyr", quiet = TRUE)
```

Install `bedtools` \
```{r engine='bash', eval=F, echo=T}
/bin/bash -c "$(curl -fsSL https://raw.githubusercontent.com/Homebrew/install/HEAD/install.sh)"
brew install bedtools
```

Install `Biostrings` \
```{r engine='R', eval=F, echo=T}
if (!requireNamespace("Biostrings", quietly = TRUE)) {
  if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
  BiocManager::install("Biostrings")
}
```

Install `latticeExtra` \
```{r engine='R', eval=TRUE, echo=T}
install.packages("latticeExtra", quiet = TRUE)
```

# Peak calling

## calling GATA peaks with `Macs3`
We want to call a set of universal peaks for the same ChIP factors. The reasons are: taking intersection of peaks between replicates/conditions will result in high false negative rates. We may lose important binding regions for transcription factor analysis. Here we are doing a compromise to call every peaks on all files and later in downstream analysis we will decide the functional peaks. \

We use `Macs3 callpeaks` to identify TF binding sites. It takes the treatment files against the control genomic input. \
`-t/--treatment FILENAME` takes your treatment file. If there is more than one alignment file, we can specify them as `-t A B C`. MACS will pool up all these files together. \
`-c/--control` takes the control file. Here we pool all the IgG file. \
`-n/--name` is the output name we give. \
`-f/--format FORMAT` we specify BAMPE format for the output. \
`-g/--gsize` is the mappable genome size or effective genome size. The actual mappable genome size is ~70% to 90% of the genome size due to the repetitive features on the chromosomes. Here we use the default hs --2.7e9 (for human genome). \
`-q/--qvalue` is the FDR cutoff to call signidicant regions. \

**GATA peaks**
```{r engine='bash', eval=F, echo=TRUE}
#! /bin/sh
#SBATCH --job-name=GATA_chip_peak_calling.sh    
#SBATCH -N 1
#SBATCH -n 1
#SBATCH -c 2
#SBATCH -p amd
#SBATCH --qos=general
#SBATCH --mem=32G
#SBATCH --mail-type=ALL
#SBATCH --mail-user=ssun@uchc.edu
#SBATCH -o chip_peak_calling.sh_%j.out
#SBATCH -e chip_peak_calling.sh_%j.err

hostname
module load macs3

mkdir temp_macs
directory=/home/FCAM/ssun/GATA3_ChIP_PRO_July2023/ChIP_final/sorted.bam_final/
# GATA_CC against IgG
macs3 callpeak --call-summits -t ${directory}*_GATA_CC*sorted.bam -c ${directory}*IgG*sorted.bam -n GATA_ChIP -g hs -q 0.01 --keep-dup all -f BAMPE --nomodel --tempdir temp_macs

#wc -l GATA_ChIP_summits.bed
#99289 
```

## Removing peaks on contigs and within blacklisted regions

Google "blacklisted genomic regions" we can find a
set of region in the genome in `bed` format that have an over-representation of
reads regardless of the experiment. We also remove peaks on non-canonical chromosomes with
`grep -v`. \

All intervals in the summits file span one base, and we use `bedtools slop` to change boundaries by adding a fixed number of bases in each direction (`-b`). These 101 base intervals will be used as input for de novo motif analysis. The 401 base intervals will be used to calculate peak intensity. The rationale for a smaller window for motif analysis is the following: \
1) the motif is typically very close to the peak summit, \
2) larger window increases the computation time substantially, and \
3) the flanking sequence is less enriched for the relevant motifs and de novo motif analysis is less sensitive with higher noise relative to background. \

We use a standard window size to calculate peak intensity, but we use a much wider window because signal 200 bases up or downstream can result from signal. We could increase the window to 600 bases and that would be fine as well. You can vary this window systematically and determine if varying the window affects the relative peak intensity rank. We use the same window size for each peak because otherwise wider windows (note the narrowPeaks file outputs are variable width) would inherently have more reads. 

```{r engine='bash', eval=F, echo=TRUE}
#! /bin/sh
#SBATCH --job-name=remove_peak.sh     
#SBATCH -N 1
#SBATCH -n 1
#SBATCH -c 2
#SBATCH -p amd
#SBATCH --qos=general
#SBATCH --mem=32G
#SBATCH --mail-type=ALL
#SBATCH --mail-user=ssun@uchc.edu
#SBATCH -o remove_peak.sh_%j.out
#SBATCH -e remove_peak.sh_%j.err
module load deeptools/3.5.0
module load bedtools
wget https://github.com/Boyle-Lab/Blacklist/raw/master/lists/hg38-blacklist.v2.bed.gz
gunzip hg38-blacklist.v2.bed.gz
blacklist=hg38-blacklist.v2.bed
sizes=/home/FCAM/ssun/Genome/hg38.chrom.sizes


for i in *_summits.bed
do
	name=$(echo $i | awk -F"/" '{print $NF}' | awk -F"_summits.bed" '{print $1}')
	echo $name
	grep -v "random" ${name}_summits.bed | grep -v "chrUn" | grep -v "chrEBV" | grep -v "chrM" | grep -v "alt" | intersectBed -v -a - -b $blacklist > ${name}_summits_final.bed
	slopBed -b 50 -i ${name}_summits_final.bed -g $sizes  | sort -k1,1 -k2,2n > ${name}_summit_100window.bed
  slopBed -b 200 -i ${name}_summits_final.bed -g $sizes  | sort -k1,1 -k2,2n > ${name}_summit_400window.bed
done
```

**peak numbers** 
```{r engine='bash', eval=F, echo=TRUE}
wc -l GATA_ChIP_summits_final.bed
#96868 
wc -l GATA_ChIP_summit_100window.bed #96868
wc -l GATA_ChIP_summit_400window.bed #96868
```


# Peak Intensity
There are two things we need to consider regarding peak intensities. The first is peak region; the second is dynamic range. Imaging a peak that is very intense but within a narrow range, the other peak is not so intense but can span its signals across a very long distance. We need to consider this region differences while calling any peak to be intense or not. \
## Counting reads in peaks (400bp window) with `DeSeq2`


```{r, engine='R', eval=F, echo=TRUE}
# mkdir GATA_Deseq2
# cd GATA_Deseq2/

# calculate the size factors 
module load samtools/1.12

dir=/home/FCAM/ssun/GATA3_ChIP_PRO_July2023/ChIP_final/sorted.bam_final/
for i in GATA
do
  echo $i
  > ${i}_header.txt
  > ${i}_reads.txt
  for j in ${dir}MCF7_dTAGGATA522*_${i}_*.sorted.bam
  do
    echo $j
    name=$(echo $j | awk -F $dir '{print $2}' | awk -F".sorted.bam" '{print $1}')
    echo $name | paste ${i}_header.txt - > ${i}_tmp.txt 
    mv ${i}_tmp.txt ${i}_header.txt
    reads=`samtools view -c -f 0x3 $j` #reads paired and mapped in proper pair
    echo $reads | paste ${i}_reads.txt - > ${i}_tmp.txt 
    mv ${i}_tmp.txt ${i}_reads.txt
  done  
  cat ${i}_header.txt ${i}_reads.txt > ${i}_tmp.txt
  mv ${i}_tmp.txt ${i}_reads.txt
  rm ${i}_header.txt
done 
```

R; load packages
```{r, engine='R', eval=F, echo=TRUE}
mkdir GATA_deseq
cd GATA_deseq
module load R/4.1.2 # this version of `R` on Xanadu has many of our libraries pre-installed. 
R 

#libraries
library(DESeq2)
library(lattice)
library(dplyr)
library(ggplot2)
library(limma)
library(bigWig)

#functions on github
source('https://raw.githubusercontent.com/mjg54/znf143_pro_seq_analysis/master/docs/ZNF143_functions.R')

#new functions
get.counts.interval <- function(df, path.to.bigWig, file.prefix = 'H') {
    vec.names = c()
    inten.df=data.frame(matrix(ncol = 0, nrow = nrow(df)))
    
    for (mod.bigWig in Sys.glob(file.path(path.to.bigWig, paste(file.prefix, "*.bigWig", sep ='')))) {
        factor.name = strsplit(strsplit(mod.bigWig, "/")[[1]][length(strsplit(mod.bigWig, "/")[[1]])], '.bigWig')[[1]][1]
        print(factor.name)
        vec.names = c(vec.names, factor.name)
        loaded.bw = load.bigWig(mod.bigWig)
        print(mod.bigWig)
        mod.inten = bed.region.bpQuery.bigWig(loaded.bw, df, abs.value = TRUE)
        inten.df = cbind(inten.df, mod.inten)
    }
    colnames(inten.df) = vec.names
    r.names = paste(df[,1], ':', df[,2], '-', df[,3], sep='')
    row.names(inten.df) = r.names
    return(inten.df)
}
```


DeSeq count read size and parse into different quantile \
```{r, engine='R', eval=F, echo=TRUE}
# recall that these summit_XXXwindow.bed has the peak region info.
a =  read.table('/home/FCAM/ssun/GATA3_ChIP_PRO_July2023/ChIP_final/peak_call/GATA_ChIP_summit_400window.bed', sep = "\t", header=FALSE)
#normalized signal
#GATA.signal.df= get.counts.interval(a, "/home/FCAM/ssun/GATA3_ChIP_PRO_July2023/ChIP_final/bigWigs/normalized_bw","MCF")
#non-normalized counts
GATA.counts.df= get.counts.interval(a, "/home/FCAM/ssun/GATA3_ChIP_PRO_July2023/ChIP_final/bigWigs/Seqoutbias_bw","MCF")
GATA.analysis.regions=GATA.counts.df[,grepl("_GATA_",colnames(GATA.counts.df))] # non-normalized counts
# identical(rownames(GATA.analysis.regions),rownames(GATA.counts.df))
# [1] TRUE

sample.conditions = factor(sapply(strsplit(colnames(GATA.analysis.regions), '_rep'), '[', 1))
deseq.counts.table = DESeqDataSetFromMatrix(countData = GATA.analysis.regions, # DESeqDataSet needs countData to be non-negative integers; non-normalized counts are integer, normalized signals has decimals.
                colData = as.data.frame(sample.conditions),
                design = ~ sample.conditions)


GATA.SF <- read.table("GATA_reads.txt", sep = '\t', header = TRUE)[,-1] # GATA size factors from read depth
GATA.size.factors = estimateSizeFactorsForMatrix(GATA.SF) # size factor transformed to smaller ratios
#MCF7_dTAGGATA522_GATA_CC_rep1 0.5385594  
#MCF7_dTAGGATA522_GATA_CC_rep2 0.5169334 
#MCF7_dTAGGATA522_GATA_CC_rep3 0.5469193  
#MCF7_dTAGGATA522_GATA_CE_rep1 0.8108480 
#MCF7_dTAGGATA522_GATA_CE_rep2 2.3452478 
#MCF7_dTAGGATA522_GATA_CE_rep3 2.1708044
                                       
sizeFactors(deseq.counts.table) <- GATA.size.factors 
# assign to each column of the count matrix (deseq.counts.table) the size factor to bring each column to a common scale
dds <- DESeq(deseq.counts.table)

normalized.counts.GATA3 = counts(dds, normalized=TRUE)
peak.intensities = rowMeans(normalized.counts.GATA3[,1:3]) # we want to get the average read counts for CC groups

quantile(peak.intensities, probs = seq(.05, 1.00, by = .05))
#         5%         10%         15%         20%         25%         30% 
#   25.03477    29.26822    32.24003    34.85009    37.44833    40.19495 
#        35%         40%         45%         50%         55%         60% 
#   43.16713    46.62380    50.27567    54.50266    59.81136    66.40026 
#        65%         70%         75%         80%         85%         90% 
#   74.63208    86.00656   102.11487   125.08599   160.58147   222.03676 
#        95%        100% 
#  348.55000 15936.61845 


names(peak.intensities) = rownames(normalized.counts.GATA3)
chr = sapply(strsplit(names(peak.intensities), ":"), "[", 1)
rnge = sapply(strsplit(names(peak.intensities), ":"), "[", 2)
start = as.numeric(sapply(strsplit(rnge, "-"), "[", 1)) + 200
end = as.numeric(sapply(strsplit(rnge, "-"), "[", 2)) - 200



# 1bp summit quantile file
j =0 
q=seq(.05, 1.00, by = .05)
count=0
for (i in quantile(peak.intensities, probs = seq(.05, 1.00, by =
.05)))
{
count = count +1

write.table(file = paste0('quantile', as.character(q[count]), '_summits.bed'), data.frame(chr[peak.intensities > j & peak.intensities <= i], start[peak.intensities > j & peak.intensities <= i], end[peak.intensities > j & peak.intensities <= i], peak.intensities[peak.intensities > j & peak.intensities <= i]), sep = '\t', quote=FALSE, col.names=FALSE, row.names=FALSE )
j = i
}

# making 100 window quantile file
start1 = as.numeric(sapply(strsplit(rnge, "-"), "[", 1)) + 150
end1 = as.numeric(sapply(strsplit(rnge, "-"), "[", 2)) - 150
j =0 
q=seq(.05, 1.00, by = .05)
count=0
for (i in quantile(peak.intensities, probs = seq(.05, 1.00, by =
.05)))
{
count = count +1

write.table(file = paste0('quantile', as.character(q[count]), '_summits_100window.bed'), data.frame(chr[peak.intensities > j & peak.intensities <= i], start1[peak.intensities > j & peak.intensities <= i], end1[peak.intensities > j & peak.intensities <= i], peak.intensities[peak.intensities > j & peak.intensities <= i]), sep = '\t', quote=FALSE, col.names=FALSE, row.names=FALSE )
j = i
}

save.image('231004_GATA3_ChIP_deseq.Rdata') 
```

```{r, engine='R', eval=F, echo=TRUE}
load('231004_GATA3_ChIP_deseq.Rdata')
options(scipen = 999)
quantile(peak.intensities, probs = seq(.05, 1.00, by = .05))
#         5%         10%         15%         20%         25%         30% 
#   25.03477    29.26822    32.24003    34.85009    37.44833    40.19495 
#        35%         40%         45%         50%         55%         60% 
#   43.16713    46.62380    50.27567    54.50266    59.81136    66.40026 
#        65%         70%         75%         80%         85%         90% 
#   74.63208    86.00656   102.11487   125.08599   160.58147   222.03676 
#        95%        100% 
#  348.55000 15936.61845 


names(peak.intensities) = rownames(normalized.counts.GATA3)
chr = sapply(strsplit(names(peak.intensities), ":"), "[", 1)
rnge = sapply(strsplit(names(peak.intensities), ":"), "[", 2)
start = as.numeric(sapply(strsplit(rnge, "-"), "[", 1)) + 200
end = as.numeric(sapply(strsplit(rnge, "-"), "[", 2)) - 200

# 1bp summit all GATA3 peaks
peak=data.frame(chr, start, end, peak.intensities)
head(peak)
#                    chr  start    end peak.intensities
#chr1:827180-827581 chr1 827380 827381         45.32469
#chr1:845596-845997 chr1 845796 845797        530.34196
#chr1:869297-869698 chr1 869497 869498         18.62632
#chr1:916569-916970 chr1 916769 916770         36.69993
#chr1:917334-917735 chr1 917534 917535        124.64502
#chr1:924653-925054 chr1 924853 924854         30.44684
nrow(peak)
#[1] 96868
ranked_peak <- peak[order(-peak[, 4]), ]
write.table(ranked_peak, file = 'GATA_ChIP_summits_final_DESeq2_ranked_intensity.bed', sep = '\t', quote=FALSE, col.names=FALSE, row.names=FALSE )
```



Now we have a ChIP-seq peak summits file that are properly ranked by its intensity calculated from 400window Deseq2. \
"GATA_ChIP_summits_final_DESeq2_ranked_intensity.bed" \
Saved in directory: /home/FCAM/ssun/GATA3_ChIP_PRO_July2023/ChIP_final/peak_Intensity/GATA_Deseq2 \


# Exhaustive MEME -MAST for GATA3

## 3-ordered Markov Background Model

GATA3 TF binding sites are short, commonly seen are 3-mer GAT/ATC or 4-mer GATA/GATC. \
To account for this fact, we want to generate the markov background model from the reference genome, which estimates the probability of a candidate motif appearing in the dataset by chance. \

I use fasta-get-markov with -m 3 option to generate a 3-ordered Markov Background Model for hg38 ref genome. The output file defines all k-mer frequencies from 1-mer to 4-mer. \

```{r, engine='bash', eval=F, echo=TRUE}
module load meme/5.4.1
module load bedtools
module load R/4.1.2

genome=/home/FCAM/ssun/Genome/hg38.fa
sizes=/home/FCAM/ssun/Genome/hg38.chrom.sizes

fasta-get-markov -m 3 $genome > hg38_bkgrnd.txt
```

```{r, engine='bash', eval=F, echo=TRUE}
head hg38_bkgrnd.txt 
# 3-order Markov frequencies from file /home/FCAM/ssun/Genome/hg38.fa
# seqs: 455    min: 970    max: 248956422    avg: 7053376.1    sum: 3209286105    alph: DNA
# order 0
A 2.950e-01
C 2.050e-01
G 2.050e-01
T 2.950e-01
# order 1
AA 9.775e-02
AC 5.051e-02


tail hg38_bkgrnd.txt 
TTCG 5.959e-04
TTCT 7.919e-03
TTGA 5.214e-03
TTGC 3.789e-03
TTGG 4.440e-03
TTGT 5.559e-03
TTTA 8.075e-03
TTTC 7.574e-03
TTTG 7.358e-03
TTTT 1.533e-02

```


<!--
## 101window

```{r, engine='bash', eval=F, echo=TRUE}
dir=/home/FCAM/ssun/GATA3_ChIP_PRO_July2023/ChIP_final/peak_Intensity/GATA_Deseq2/
head ${dir}GATA_ChIP_summits_final_DESeq2_ranked_intensity.bed
chr17	61231384	61231385	15936.6184521741
chr20	57264400	57264401	12166.7947008285
chr20	54456658	54456659	11926.1802605615
chr17	61219540	61219541	10340.3616996713
chr20	50764408	50764409	8164.47826610042
chr17	61236076	61236077	7391.51505715346
chr20	53774104	53774105	7166.66522528761
chr17	61725015	61725016	6641.5850441495
chr20	53786233	53786234	6270.21535416316
chr20	53786498	53786499	6051.79992360185

```

(cd GATA3_ChIP_PRO_July2023/ChIP_final/new_Exhaustive_MEME_MAST_round1to8_240304/) \

```{r engine='bash', eval=F, echo=TRUE}
module load bedtools
sizes=/home/FCAM/ssun/Genome/hg38.chrom.sizes
genome=/home/FCAM/ssun/Genome/hg38.fa
dir=/home/FCAM/ssun/GATA3_ChIP_PRO_July2023/ChIP_final/peak_Intensity/GATA_Deseq2/

slopBed -b 50 -i ${dir}GATA_ChIP_summits_final_DESeq2_ranked_intensity.bed -g $sizes  | sort -k1,1 -k2,2n > GATA_ChIP_summits_final_DESeq2_ranked_intensity_101bpwindow.bed #96868

fastaFromBed -fi $genome -bed GATA_ChIP_summits_final_DESeq2_ranked_intensity_101bpwindow.bed -fo GATA_ChIP_summits_final_DESeq2_ranked_intensity_101bpwindow.fasta

```

-->


## 161window

(cd GATA3_ChIP_PRO_July2023/ChIP_final/new_Exhaustive_MEME_MAST_round1to8_240304/) \
```{r engine='bash', eval=F, echo=TRUE}
module load bedtools
sizes=/home/FCAM/ssun/Genome/hg38.chrom.sizes
genome=/home/FCAM/ssun/Genome/hg38.fa
dir=/home/FCAM/ssun/GATA3_ChIP_PRO_July2023/ChIP_final/peak_Intensity/GATA_Deseq2/

slopBed -b 80 -i ${dir}GATA_ChIP_summits_final_DESeq2_ranked_intensity.bed -g $sizes  | sort -k1,1 -k2,2n > GATA_ChIP_summits_final_DESeq2_ranked_intensity_161bpwindow.bed #96868

fastaFromBed -fi $genome -bed GATA_ChIP_summits_final_DESeq2_ranked_intensity_161bpwindow.bed -fo GATA_ChIP_summits_final_DESeq2_ranked_intensity_161bpwindow.fasta

```


## MEME-round1:


Sort peaks by intensity, and select top 5000 peaks. \
```{r engine='bash', eval=F, echo=TRUE}
# sort out the top 10000 GATA peaks and search for top 1 motif
sort -nrk4,4 GATA_ChIP_summits_final_DESeq2_ranked_intensity_161bpwindow.bed | head -n 5000 > GATA_ChIP_top5000_161window.bed #5000
```

Convert this top 5000 peak bed file to .fasta file with `fastaFromBed`. \
```{r engine='bash', eval=F, echo=TRUE}
fastaFromBed -fi $genome -bed GATA_ChIP_top5000_161window.bed -fo GATA_ChIP_top5000_161window.fasta
```

Perform MEME on top 5000 GATA3 peaks with -minw 4, -maxw 35 and the 3-ordered Markov Background Model. \
**And the default wg/ws setting**: \
```{r engine='bash', eval=F, echo=TRUE}
#!/bin/bash
#SBATCH --job-name=r1_meme_top5000.sh
#SBATCH -N 6
#SBATCH -n 6
#SBATCH -c 12
#SBATCH -p general
#SBATCH --qos=general
#SBATCH --mem=80G
#SBATCH --mail-type=ALL
#SBATCH --mail-user=ssun@uchc.edu
#SBATCH -o r1_meme_top5000.sh_%j.out
#SBATCH -e r1_meme_top5000.sh_%j.err

hostname
export TMPDIR="${HOME}/temp"
module load meme/5.4.1

meme -p 72 GATA_ChIP_top5000_161window.fasta -oc GATA_ChIP_top5000_161window_meme_output -nmotifs 10 -csites 20000 -objfun classic -searchsize 0 -minw 4 -maxw 35 -revcomp -dna -bfile hg38_bkgrnd.txt -maxsize 10000000
```

MEME identifies 4 GATA3-like motifs. \

**A small wg/ws setting:** \

```{r engine='bash', eval=F, echo=TRUE}
#!/bin/bash
#SBATCH --job-name=r1ws1_meme_top5000.sh
#SBATCH -N 6
#SBATCH -n 6
#SBATCH -c 12
#SBATCH -p general
#SBATCH --qos=general
#SBATCH --mem=80G
#SBATCH --mail-type=ALL
#SBATCH --mail-user=ssun@uchc.edu
#SBATCH -o r1ws1_meme_top5000.sh_%j.out
#SBATCH -e r1ws1_meme_top5000.sh_%j.err

hostname
export TMPDIR="${HOME}/temp"
module load meme/5.4.1

meme -p 72 GATA_ChIP_top5000_161window.fasta -oc GATA_ChIP_top5000_161window_ws1_meme_output -nmotifs 10 -csites 20000 -objfun classic -searchsize 0 -minw 4 -maxw 35 -revcomp -dna -bfile hg38_bkgrnd.txt -maxsize 10000000 -wg 1 -ws 1
```

The small gap-open panalty parameter `-ws==1` did not generate meaningful results. \

### MAST (p<0.0005)

Here we are performing MAST to find peaks with the top GATA3-like motifs of the four. \
**Extract MEME PSWMs** \
First isolate the PSWMs associated with GATA-like motifs. There are four GATA-like motifs in first round MEME: \
```{r engine='bash', eval=F, echo=TRUE}
wget https://raw.githubusercontent.com/sysunn/siyu_daily_update/main/December_2023/MEME_individual_from_db_python3.py
python MEME_individual_from_db_python3.py -i GATA_ChIP_top5000_161window_meme_output/meme.txt
#WGATAAVATCW_meme.txt -- top enriched GATA-like motif in round1
#GATAADVATCW_meme.txt
#HTTATCTNYHHATCT_meme.txt
#ATCWGATAAVN_meme.txt


#mkdir individual_meme_round1
cp WGATAAVATCW_meme.txt individual_meme_round1/GATAmotif1_1_meme.txt
cp GATAADVATCW_meme.txt individual_meme_round1/GATAmotif1_2_meme.txt
cp HTTATCTNYHHATCT_meme.txt individual_meme_round1/GATAmotif1_3_meme.txt
cp ATCWGATAAVN_meme.txt individual_meme_round1/GATAmotif1_4_meme.txt
rm *meme.txt
```

We will MAST only the top enriched motif, with a P-value<0.0005. \
```{r engine='bash', eval=F, echo=TRUE}
cat individual_meme_round1/GATAmotif1_1_meme.txt 
MEME version 5.4.1 (Release date: Sat Aug 21 19:23:23 2021 -0700)

ALPHABET= ACGT

strands: + -

A 0.295 C 0.205 G 0.205 T 0.295

MOTIF	WGATAAVATCW
letter-probability matrix: alength= 4 w= 11 nsites= 1090 E= 3.8e-740 
 0.600000  0.044954  0.046789  0.308257 
 0.000000  0.000000  1.000000  0.000000 
 1.000000  0.000000  0.000000  0.000000 
 0.000000  0.000000  0.000000  1.000000 
 0.837615  0.000000  0.000000  0.162385 
 0.555046  0.131193  0.185321  0.128440 
 0.273394  0.363303  0.238532  0.124771 
 1.000000  0.000000  0.000000  0.000000 
 0.000000  0.000000  0.000000  1.000000 
 0.000000  1.000000  0.000000  0.000000 
 0.433945  0.000917  0.023853  0.541284 
```

```{r engine='bash', eval=F, echo=TRUE}
module load meme/5.4.1
module load R/4.1.2
mast -mt 0.0005 -hit_list -best individual_meme_round1/GATAmotif1_1_meme.txt GATA_ChIP_summits_final_DESeq2_ranked_intensity_161bpwindow.fasta > mast_GATA3_PSWM_in_peaks_round1.txt #27756
Rscript /home/FCAM/ssun/scripts/parse_mast_to_coordinates.R mast_GATA3_PSWM_in_peaks_round1.txt

wc -l mast_GATA3_PSWM_in_peaks_round1.bed 
#27753
```

Use `intersectBed` to find GATA3 peaks without motif 1. \
And convert to .fasta \
```{r engine='bash', eval=F, echo=TRUE}
module load bedtools
genome=/home/FCAM/ssun/Genome/hg38.fa
# From all peaks exclude peaks with motif1 
intersectBed -v -a GATA_ChIP_summits_final_DESeq2_ranked_intensity_161bpwindow.bed -b mast_GATA3_PSWM_in_peaks_round1.bed > without_motifs_1.bed #69112
fastaFromBed -fi $genome -bed without_motifs_1.bed -fo without_motifs_1.fasta
```



## MEME-round2:
sort and select top 5000 from peaks without motif 1. \
```{r engine='bash', eval=F, echo=TRUE}
# select top 5000
sort -nrk4,4 without_motifs_1.bed | head -n 5000 > without_motifs_1_top5000.bed
fastaFromBed -fi $genome -bed without_motifs_1_top5000.bed -fo without_motifs_1_top5000.fasta
```

Perform MEME on top 5000 GATA3 peaks without motif1 with -minw 4, -maxw 35 and the 3-ordered Markov Background Model. \
**And the default wg/ws setting**: \
```{r engine='bash', eval=F, echo=TRUE}
#!/bin/bash
#SBATCH --job-name=r2_meme_top5000.sh
#SBATCH -N 6
#SBATCH -n 6
#SBATCH -c 12
#SBATCH -p general
#SBATCH --qos=general
#SBATCH --mem=80G
#SBATCH --mail-type=ALL
#SBATCH --mail-user=ssun@uchc.edu
#SBATCH -o r2_meme_top5000.sh_%j.out
#SBATCH -e r2_meme_top5000.sh_%j.err

hostname
export TMPDIR="${HOME}/temp"
module load meme/5.4.1

meme -p 72 without_motifs_1_top5000.fasta -oc GATA_ChIP_without_motifs_1_top5000_161window_meme_output -nmotifs 10 -csites 20000 -objfun classic -searchsize 0 -minw 4 -maxw 35 -revcomp -dna -bfile hg38_bkgrnd.txt -maxsize 10000000
```

This time, the top enriched motif is the second GATA-like motif in the first round. \

**A small wg/ws setting:** \

```{r engine='bash', eval=F, echo=TRUE}
#!/bin/bash
#SBATCH --job-name=r2ws1_meme_top5000.sh
#SBATCH -N 6
#SBATCH -n 6
#SBATCH -c 12
#SBATCH -p general
#SBATCH --qos=general
#SBATCH --mem=80G
#SBATCH --mail-type=ALL
#SBATCH --mail-user=ssun@uchc.edu
#SBATCH -o r2ws1_meme_top5000.sh_%j.out
#SBATCH -e r2ws1_meme_top5000.sh_%j.err

hostname
export TMPDIR="${HOME}/temp"
module load meme/5.4.1

meme -p 72 without_motifs_1_top5000.fasta -oc GATA_ChIP_without_motifs_1_top5000_161window_ws1_meme_output -nmotifs 10 -csites 20000 -objfun classic -searchsize 0 -minw 4 -maxw 35 -revcomp -dna -bfile hg38_bkgrnd.txt -maxsize 10000000 -wg 1 -ws 1
```

### MAST (p<0.0005)

Here we are performing MAST to find peaks with the top GATA3-like motifs of the three. \
**Extract MEME PSWMs** \
First isolate the PSWMs associated with GATA-like motifs. There are four GATA-like motifs in first round MEME: \
```{r engine='bash', eval=F, echo=TRUE}
#wget https://raw.githubusercontent.com/sysunn/siyu_daily_update/main/December_2023/MEME_individual_from_db_python3.py
python MEME_individual_from_db_python3.py -i GATA_ChIP_without_motifs_1_top5000_161window_meme_output/meme.txt
#AGATBYTTATC_meme.txt -- top enriched GATA-like motif in round2


mkdir individual_meme
cp individual_meme_round1/GATAmotif1_1_meme.txt individual_meme/GATAmotif1_meme.txt # save round1 motif to this folder
cp AGATBYTTATC_meme.txt individual_meme/GATAmotif2_meme.txt
rm *meme.txt
```

We will MAST only the top enriched motif, with a P-value<0.0005. \
```{r engine='bash', eval=F, echo=TRUE}
cat individual_meme/GATAmotif2_meme.txt 
MEME version 5.4.1 (Release date: Sat Aug 21 19:23:23 2021 -0700)

ALPHABET= ACGT

strands: + -

A 0.295 C 0.205 G 0.205 T 0.295

MOTIF	AGATBYTTATC
letter-probability matrix: alength= 4 w= 11 nsites= 1080 E= 3.9e-629 
 0.652778  0.026852  0.000000  0.320370 
 0.000000  0.000000  1.000000  0.000000 
 0.999074  0.000926  0.000000  0.000000 
 0.000000  0.000000  0.000000  1.000000 
 0.092593  0.261111  0.248148  0.398148 
 0.180556  0.250000  0.179630  0.389815 
 0.069444  0.190741  0.086111  0.653704 
 0.224074  0.000000  0.019444  0.756481 
 1.000000  0.000000  0.000000  0.000000 
 0.000000  0.000000  0.000000  1.000000 
 0.000000  1.000000  0.000000  0.000000
```

```{r engine='bash', eval=F, echo=TRUE}
module load meme/5.4.1
module load R/4.1.2
mast -mt 0.0005 -hit_list -best individual_meme/GATAmotif2_meme.txt without_motifs_1.fasta > mast_GATA3_PSWM_in_peaks_round2.txt #16778
Rscript /home/FCAM/ssun/scripts/parse_mast_to_coordinates.R mast_GATA3_PSWM_in_peaks_round2.txt

wc -l mast_GATA3_PSWM_in_peaks_round2.bed 
#16775
```

Use `intersectBed` to find GATA3 peaks without motif 1. \
And convert to .fasta \
```{r engine='bash', eval=F, echo=TRUE}
module load bedtools
genome=/home/FCAM/ssun/Genome/hg38.fa
# From peaks without motif1 exclude peaks with motif2
intersectBed -v -a without_motifs_1.bed -b mast_GATA3_PSWM_in_peaks_round2.bed > without_motifs_12.bed #52334
fastaFromBed -fi $genome -bed without_motifs_12.bed -fo without_motifs_12.fasta
```

## MEME-round3:
sort and select top 5000 from peaks without motif 1 and motif2. \
```{r engine='bash', eval=F, echo=TRUE}
# select top 5000
sort -nrk4,4 without_motifs_12.bed | head -n 5000 > without_motifs_12_top5000.bed
fastaFromBed -fi $genome -bed without_motifs_12_top5000.bed -fo without_motifs_12_top5000.fasta
```

Perform MEME on top 5000 GATA3 peaks without motif1 and motif2 with -minw 4, -maxw 35 and the 3-ordered Markov Background Model. \
**And the default wg/ws setting**: \
```{r engine='bash', eval=F, echo=TRUE}
#!/bin/bash
#SBATCH --job-name=r3_meme_top5000.sh
#SBATCH -N 6
#SBATCH -n 6
#SBATCH -c 12
#SBATCH -p general
#SBATCH --qos=general
#SBATCH --mem=80G
#SBATCH --mail-type=ALL
#SBATCH --mail-user=ssun@uchc.edu
#SBATCH -o r3_meme_top5000.sh_%j.out
#SBATCH -e r3_meme_top5000.sh_%j.err

hostname
export TMPDIR="${HOME}/temp"
module load meme/5.4.1

meme -p 72 without_motifs_12_top5000.fasta -oc GATA_ChIP_without_motifs_12_top5000_161window_meme_output -nmotifs 10 -csites 20000 -objfun classic -searchsize 0 -minw 4 -maxw 35 -revcomp -dna -bfile hg38_bkgrnd.txt -maxsize 10000000
```



The second motif is the GATA-like motif, two GATA sites with 5 spacings. \



### MAST (p<0.0005)

Here we are performing MAST to find peaks with the top GATA3-like motifs of the three. \
**Extract MEME PSWMs** \
First isolate the PSWMs associated with GATA-like motifs. There are four GATA-like motifs in first round MEME: \
```{r engine='bash', eval=F, echo=TRUE}
#wget https://raw.githubusercontent.com/sysunn/siyu_daily_update/main/December_2023/MEME_individual_from_db_python3.py
python MEME_individual_from_db_python3.py -i GATA_ChIP_without_motifs_12_top5000_161window_meme_output/meme.txt
#AGATDDDNAGATAAD_meme.txt -- top enriched GATA-like motif in round3

cp AGATDDDNAGATAAD_meme.txt individual_meme/GATAmotif3_meme.txt
rm *meme.txt
```

We will MAST only the top enriched motif, with a P-value<0.0005. \
```{r engine='bash', eval=F, echo=TRUE}
module load meme/5.4.1
module load R/4.1.2
mast -mt 0.0005 -hit_list -best individual_meme/GATAmotif3_meme.txt without_motifs_12.fasta > mast_GATA3_PSWM_in_peaks_round3.txt #9430
Rscript /home/FCAM/ssun/scripts/parse_mast_to_coordinates.R mast_GATA3_PSWM_in_peaks_round3.txt

wc -l mast_GATA3_PSWM_in_peaks_round3.bed 
#9427
```

Use `intersectBed` to find GATA3 peaks without motif 1. \
And convert to .fasta \
```{r engine='bash', eval=F, echo=TRUE}
module load bedtools
genome=/home/FCAM/ssun/Genome/hg38.fa
# From peaks without motif1 exclude peaks with motif2
intersectBed -v -a without_motifs_12.bed -b mast_GATA3_PSWM_in_peaks_round3.bed > without_motifs_123.bed #42904
fastaFromBed -fi $genome -bed without_motifs_123.bed -fo without_motifs_123.fasta
```


## MEME-round4:

sort and select top 5000 from peaks without motif 1, 2, 3. \
```{r engine='bash', eval=F, echo=TRUE}
# select top 5000
sort -nrk4,4 without_motifs_123.bed | head -n 5000 > without_motifs_123_top5000.bed
fastaFromBed -fi $genome -bed without_motifs_123_top5000.bed -fo without_motifs_123_top5000.fasta
```

Perform MEME on top 5000 GATA3 peaks without motif123 with -minw 4, -maxw 35 and the 3-ordered Markov Background Model. \
**And the default wg/ws setting**: \
```{r engine='bash', eval=F, echo=TRUE}
#!/bin/bash
#SBATCH --job-name=r4_meme_top5000.sh
#SBATCH -N 6
#SBATCH -n 6
#SBATCH -c 12
#SBATCH -p general
#SBATCH --qos=general
#SBATCH --mem=80G
#SBATCH --mail-type=ALL
#SBATCH --mail-user=ssun@uchc.edu
#SBATCH -o r4_meme_top5000.sh_%j.out
#SBATCH -e r4_meme_top5000.sh_%j.err

hostname
export TMPDIR="${HOME}/temp"
module load meme/5.4.1

meme -p 72 without_motifs_123_top5000.fasta -oc GATA_ChIP_without_motifs_123_top5000_161window_meme_output -nmotifs 10 -csites 20000 -objfun classic -searchsize 0 -minw 4 -maxw 35 -revcomp -dna -bfile hg38_bkgrnd.txt -maxsize 10000000
```

The ninth motif is a GATA-like motif which is GAT-GAT with 6 spacings. \

### MAST (p<0.0005)

Here we are performing MAST to find peaks with the top GATA3-like motifs of the three. \
**Extract MEME PSWMs** \
First isolate the PSWMs associated with GATA-like motifs. There are four GATA-like motifs in first round MEME: \
```{r engine='bash', eval=F, echo=TRUE}
#wget https://raw.githubusercontent.com/sysunn/siyu_daily_update/main/December_2023/MEME_individual_from_db_python3.py
python MEME_individual_from_db_python3.py -i GATA_ChIP_without_motifs_123_top5000_161window_meme_output/meme.txt
#TTATCTBYNNVATCT_meme.txt -- top enriched GATA-like motif in round4

cp TTATCTBYNNVATCT_meme.txt individual_meme/GATAmotif4_meme.txt
rm *meme.txt
```

We will MAST only the top enriched motif, with a P-value<0.0005. \
```{r engine='bash', eval=F, echo=TRUE}
module load meme/5.4.1
module load R/4.1.2
mast -mt 0.0005 -hit_list -best individual_meme/GATAmotif4_meme.txt without_motifs_123.fasta > mast_GATA3_PSWM_in_peaks_round4.txt #7434
Rscript /home/FCAM/ssun/scripts/parse_mast_to_coordinates.R mast_GATA3_PSWM_in_peaks_round4.txt

wc -l mast_GATA3_PSWM_in_peaks_round4.bed 
#7431
```

Use `intersectBed` to find GATA3 peaks without motif 1. \
And convert to .fasta \
```{r engine='bash', eval=F, echo=TRUE}
module load bedtools
genome=/home/FCAM/ssun/Genome/hg38.fa
# From peaks without motif1 exclude peaks with motif2
intersectBed -v -a without_motifs_123.bed -b mast_GATA3_PSWM_in_peaks_round4.bed > without_motifs_1234.bed #35473
fastaFromBed -fi $genome -bed without_motifs_1234.bed -fo without_motifs_1234.fasta
```

## MEME-round5:
sort and select top 5000 from peaks without motif 1, 2, 3, 4. \
```{r engine='bash', eval=F, echo=TRUE}
# select top 5000
sort -nrk4,4 without_motifs_1234.bed | head -n 5000 > without_motifs_1234_top5000.bed
fastaFromBed -fi $genome -bed without_motifs_1234_top5000.bed -fo without_motifs_1234_top5000.fasta
```

Perform MEME on top 5000 GATA3 peaks without motif1234 with -minw 4, -maxw 35 and the 3-ordered Markov Background Model. \

**And the default wg/ws setting**: \
```{r engine='bash', eval=F, echo=TRUE}
#!/bin/bash
#SBATCH --job-name=r5_meme_top5000.sh
#SBATCH -N 6
#SBATCH -n 6
#SBATCH -c 12
#SBATCH -p general
#SBATCH --qos=general
#SBATCH --mem=80G
#SBATCH --mail-type=ALL
#SBATCH --mail-user=ssun@uchc.edu
#SBATCH -o r5_meme_top5000.sh_%j.out
#SBATCH -e r5_meme_top5000.sh_%j.err

hostname
export TMPDIR="${HOME}/temp"
module load meme/5.4.1

meme -p 72 without_motifs_1234_top5000.fasta -oc GATA_ChIP_without_motifs_1234_top5000_161window_meme_output -nmotifs 10 -csites 20000 -objfun classic -searchsize 0 -minw 4 -maxw 35 -revcomp -dna -bfile hg38_bkgrnd.txt -maxsize 10000000
```

**default wg/ws setting, search for 20 motifs**: \
```{r engine='bash', eval=F, echo=TRUE}
#!/bin/bash
#SBATCH --job-name=r51_meme_top5000.sh
#SBATCH -N 6
#SBATCH -n 6
#SBATCH -c 12
#SBATCH -p general
#SBATCH --qos=general
#SBATCH --mem=80G
#SBATCH --mail-type=ALL
#SBATCH --mail-user=ssun@uchc.edu
#SBATCH -o r51_meme_top5000.sh_%j.out
#SBATCH -e r51_meme_top5000.sh_%j.err

hostname
export TMPDIR="${HOME}/temp"
module load meme/5.4.1

meme -p 72 without_motifs_1234_top5000.fasta -oc GATA_ChIP_without_motifs_1234_top5000_161window_20_meme_output -nmotifs 20 -csites 20000 -objfun classic -searchsize 0 -minw 4 -maxw 35 -revcomp -dna -bfile hg38_bkgrnd.txt -maxsize 10000000
```


**small wg/ws setting**: \
```{r engine='bash', eval=F, echo=TRUE}
#!/bin/bash
#SBATCH --job-name=r5ws1_meme_top5000.sh
#SBATCH -N 6
#SBATCH -n 6
#SBATCH -c 12
#SBATCH -p general
#SBATCH --qos=general
#SBATCH --mem=80G
#SBATCH --mail-type=ALL
#SBATCH --mail-user=ssun@uchc.edu
#SBATCH -o r5ws1_meme_top5000.sh_%j.out
#SBATCH -e r5ws1_meme_top5000.sh_%j.err

hostname
export TMPDIR="${HOME}/temp"
module load meme/5.4.1

meme -p 72 without_motifs_1234_top5000.fasta -oc GATA_ChIP_without_motifs_1234_top5000_161window_ws1_meme_output -nmotifs 10 -csites 20000 -objfun classic -searchsize 0 -minw 10 -maxw 35 -revcomp -dna -bfile hg38_bkgrnd.txt -maxsize 10000000 -wg 1 -ws 1
```

The small wg/ws did not generate useful motifs, but the default MEME search for 20 motifs provided the 14th motif, a GATA-like motif: GAT-ATC with 1 spacing. Its E-value is 1.2e-065, indicating it is still a significant motif. \



### MAST (p<0.0005)

Here we are performing MAST to find peaks with the top GATA3-like motifs of the three. \
**Extract MEME PSWMs** \
First isolate the PSWMs associated with GATA-like motifs. There are four GATA-like motifs in first round MEME: \
```{r engine='bash', eval=F, echo=TRUE}
#wget https://raw.githubusercontent.com/sysunn/siyu_daily_update/main/December_2023/MEME_individual_from_db_python3.py
python MEME_individual_from_db_python3.py -i GATA_ChIP_without_motifs_1234_top5000_161window_20_meme_output/meme.txt
#NCTTATCWGAT_meme.txt -- top enriched GATA-like motif in round5

cp NCTTATCWGAT_meme.txt individual_meme/GATAmotif5_meme.txt
rm *meme.txt
```

We will MAST only the top enriched motif, with a P-value<0.0005. \
```{r engine='bash', eval=F, echo=TRUE}
module load meme/5.4.1
module load R/4.1.2
mast -mt 0.0005 -hit_list -best individual_meme/GATAmotif5_meme.txt without_motifs_1234.fasta > mast_GATA3_PSWM_in_peaks_round5.txt #4654
Rscript /home/FCAM/ssun/scripts/parse_mast_to_coordinates.R mast_GATA3_PSWM_in_peaks_round5.txt

wc -l mast_GATA3_PSWM_in_peaks_round5.bed 
#4651
```

Use `intersectBed` to find GATA3 peaks without motif 12345. \
And convert to .fasta \
```{r engine='bash', eval=F, echo=TRUE}
module load bedtools
genome=/home/FCAM/ssun/Genome/hg38.fa
# From peaks without motif1234 exclude peaks with motif
intersectBed -v -a without_motifs_1234.bed -b mast_GATA3_PSWM_in_peaks_round5.bed > without_motifs_12345.bed #30822
fastaFromBed -fi $genome -bed without_motifs_12345.bed -fo without_motifs_12345.fasta
```

## MEME-round6:

sort and select top 5000 from peaks without motif 1, 2, 3, 4, 5. \
```{r engine='bash', eval=F, echo=TRUE}
# select top 5000
sort -nrk4,4 without_motifs_12345.bed | head -n 5000 > without_motifs_12345_top5000.bed
fastaFromBed -fi $genome -bed without_motifs_12345_top5000.bed -fo without_motifs_12345_top5000.fasta
```

Perform MEME on top 5000 GATA3 peaks without motif12345 with -minw 4, -maxw 35 and the 3-ordered Markov Background Model. \

**And the default wg/ws setting**: \
```{r engine='bash', eval=F, echo=TRUE}
#!/bin/bash
#SBATCH --job-name=r6_meme_top5000.sh
#SBATCH -N 6
#SBATCH -n 6
#SBATCH -c 12
#SBATCH -p general
#SBATCH --qos=general
#SBATCH --mem=80G
#SBATCH --mail-type=ALL
#SBATCH --mail-user=ssun@uchc.edu
#SBATCH -o r6_meme_top5000.sh_%j.out
#SBATCH -e r6_meme_top5000.sh_%j.err

hostname
export TMPDIR="${HOME}/temp"
module load meme/5.4.1

meme -p 72 without_motifs_12345_top5000.fasta -oc GATA_ChIP_without_motifs_12345_top5000_161window_20_meme_output -nmotifs 20 -csites 20000 -objfun classic -searchsize 0 -minw 4 -maxw 35 -revcomp -dna -bfile hg38_bkgrnd.txt -maxsize 10000000
```


```{r engine='bash', eval=F, echo=TRUE}
#!/bin/bash
#SBATCH --job-name=r6ws1_meme_top5000.sh
#SBATCH -N 6
#SBATCH -n 6
#SBATCH -c 12
#SBATCH -p general
#SBATCH --qos=general
#SBATCH --mem=80G
#SBATCH --mail-type=ALL
#SBATCH --mail-user=ssun@uchc.edu
#SBATCH -o r6ws1_meme_top5000.sh_%j.out
#SBATCH -e r6ws1_meme_top5000.sh_%j.err

hostname
export TMPDIR="${HOME}/temp"
module load meme/5.4.1

meme -p 72 without_motifs_12345_top5000.fasta -oc GATA_ChIP_without_motifs_12345_top5000_161window_20ws1_meme_output -nmotifs 20 -csites 20000 -objfun classic -searchsize 0 -minw 4 -maxw 35 -revcomp -dna -bfile hg38_bkgrnd.txt -maxsize 10000000 -wg 1 -ws 1
```


## MEME-round7:
## MEME-round8:


