---
title: \sf GATA3_ChIP_T2T
header-includes:
- \usepackage{color}
- \usepackage{float}
- \DeclareUnicodeCharacter{2212}{-}
date: "February 20, 2024"
output:
  bookdown::html_document2:
    toc: true
fontsize: 14pt
geometry: margin=1in
---

```{css, echo=FALSE}
body .main-container {
  max-width: 1200px !important;
  width: 1200px !important;
}
body {
  max-width: 1200px !important;
}

pre {
  max-height: 600px;
  overflow-y: auto;
}

pre[class] {
  max-height: 600px;
}
```

```{css, echo=FALSE}
.watch-out {
  background-color: lightcyan;
  border: 3px solid lightgrey;
  font-weight: bold;
}

.revision {
  background-color: #FFF4F2;
  border: 3px solid lightgrey;
  font-weight: bold;
}

.highlighted
{
   color:red;
}

.highlighted2
{
   color:blue;
}
```

```{r engine='R', eval=TRUE, echo=F}
knitr::opts_chunk$set(class.source = "watch-out")
```


# Install required packages

# GATA3 ChIP-seq

## Align to the T2T human genome

First we want to generate a folder in the convenient directory and name it as Genome. We will save all the genome files here. `cd` to this working directory, we are getting reference genome and chrom.sizes file from the USCS genome server, and build the genome index with `bowtie2-build`.

```{r engine='bash', eval=F, echo=TRUE}
module load genometools/1.5.10
module load bowtie2
export PATH=$PATH:/home/FCAM/ssun/packages

#cd /home/FCAM/ssun/Genome/
wget https://hgdownload.soe.ucsc.edu/goldenPath/hs1/bigZips/hs1.chrom.sizes.txt 
#wget https://hgdownload.soe.ucsc.edu/goldenPath/hs1/bigZips/hs1.fa.gz
#gunzip hs1.fa.gz
cd hs1_bt2
bowtie2-build ../hs1.fa hs1

```

Now we are aligning the processed fastq files (no_adapt.fastq) to the T2T reference genome: `hs1.fa`. \
I am only processing the three GATA_CC libraries and the four IgG libraries. \

```{r engine='bash', eval=F, echo=TRUE}
#cd /home/FCAM/ssun/GATA3_ChIP_PRO_July2023/ChIP_final/ChIP_GATA3_CC_IgG_MCF7_align_to_T2T
cp ../ChIP1_Sathyan_MCF7_GATA3/1_DataProcessing_\&Alignment/ChIP1_Sathyan_MCF7_GATA3/MCF7_dTAGGATA522_GATA_CC_rep*_PE*_no_adapt.fastq.gz .
cp ../ChIP1_Sathyan_MCF7_GATA3/1_DataProcessing_\&Alignment/ChIP1_Sathyan_MCF7_GATA3/MCF7_dTAGGATA522_IgG_*_no_adapt.fastq.gz .
cp ../ChIP2_Sathyan_MCF7_GATA3/no_adapt/MCF7_dTAGGATA522_IgG_CC_rep4_PE*_no_adapt.fastq.gz .

gunzip *.gz
```


T2T_chip_alignment_240220.sh  \
```{r engine='bash', eval=F, echo=TRUE}
#!/bin/bash
#SBATCH --job-name=T2T_chip_alignment_240220.sh     
#SBATCH -N 1                  
#SBATCH -n 1                 
#SBATCH -c 32                  
#SBATCH -p amd           
#SBATCH --qos=general       
#SBATCH --mem=32G               
#SBATCH --mail-type=ALL 
#SBATCH --mail-user=ssun@uchc.edu
#SBATCH -o T2T_chip_alignment_240220.sh_%j.out
#SBATCH -e T2T_chip_alignment_240220.sh_%j.err

hostname
name=240220

# I have seqOutBias and other packages export to my Path
export PATH=$PATH:/home/FCAM/ssun/packages/

module load samtools/1.16.1
module load genometools/1.5.10
module load ucsc_genome/2012.05.22
module load rust
module load bowtie2
module load bedtools

export TMPDIR="${HOME}/temp" 
sizes=/home/FCAM/ssun/Genome/hs1.chrom.sizes.txt

genome=/home/FCAM/ssun/Genome/hs1.fa
genome_index=/home/FCAM/ssun/Genome/hs1_bt2/hs1
ncore=32 
tallymer=hs1.tal_42.gtTxt.gz

#gzip -d ${name}_PE1_no_adapt.fastq.gz
#gzip -d ${name}_PE2_no_adapt.fastq.gz
bowtie2 -p $ncore --maxins 800 -x $genome_index -1 ${name}_PE1_no_adapt.fastq -2 ${name}_PE2_no_adapt.fastq | samtools sort -@ $ncore -n -o ${name}.bw.bam
gzip ${name}_PE1_no_adapt.fastq
gzip ${name}_PE2_no_adapt.fastq
#gzip ${name}_PE2.fastq
#gzip ${name}_PE1.fastq
samtools fixmate -m ${name}.bw.bam - | samtools sort -@ $ncore - | samtools markdup -s -r - ${name}.hs1.bam
seqOutBias ${genome} ${name}.hs1.bam --shift-counts --no-scale \
                                      --bw=${name}.bigWig --read-size=42 --tallymer=$tallymer 2>&1 | tee ${name}_seqOutBias.log
samtools sort -@ $ncore -n -o ${name}.sorted.bam ${name}.hs1.bam
# the above samtools -n is sorting files by queryname, which is important for the next bedtools bamtobed command with the -bedpe flag; 
# the bedtools bamtobed -bedpe is generating warning signs:
# *****WARNING: Query {...} is marked as paired, but its mate does not occur next to it in your BAM file.  Skipping.
# This might due to some single end sequences, I checked the fraction of these not paired sequences, it is <1.5%, so we don't need to worry.
bedtools bamtobed -i ${name}.sorted.bam -bedpe > ${name}_bed12.bed

awk '$1==$4 {print $0}' ${name}_bed12.bed | awk '{OFS="\t";} {print $1, $2, $6}' | awk '$1!="." && $3>$2 && (($3 - $2)<2000) {print $0}' | sort -k1,1 -k2,2n > ${name}_read_span.bed
genomeCoverageBed -bg -i ${name}_read_span.bed -g $sizes > ${name}.bedGraph
depth=`awk -F'\t' '{sum+=$5;}END{print sum;}' ${name}.hs1_not_scaled.bed`
echo "depth is " $depth
scaled=$(bc <<< "scale=3 ; 10000000 / $depth")
echo "scale is " $scaled
awk -v scaled="$scaled" '{OFS="\t";} {print $1, $2, $3, $4*scaled}' ${name}.bedGraph > ${name}_normalized.bedGraph
wigToBigWig -clip ${name}_normalized.bedGraph $sizes ${name}_normalized.bigWig
echo "complete"
```

**Run the previous chunk in parallel**

Now we are generating tallymer files and table use `seqOutBias` outside the loop, and then run the previous chunk in parallel. \
We need to know the read size of our library. Here the read size is 42.\
```{r engine='bash', eval=F, echo=TRUE}
#Compute mappability for the given read length and the k-mer that corresponds to each possible read alignment position
#time-consuming but only need to run this one time
seqOutBias seqtable hs1.fa --read-size=42
mkdir -p ${HOME}/temp

file=T2T_chip_alignment_240220.sh  

for i in *_PE1_no_adapt.fastq 
do
    nm=$(echo $i | awk -F"/" '{print $NF}' | awk -F"_PE1_no_adapt.fastq" '{print $1}')
    echo $nm
    sed -e "s/240220/${nm}/g" "$file" > pro_processing_${nm}.sh
    sbatch pro_processing_${nm}.sh
    sleep 1
done
```

# Peak calling

## calling GATA peaks with `Macs3`

**GATA peaks**
```{r engine='bash', eval=F, echo=TRUE}
#! /bin/sh
#SBATCH --job-name=GATA_chip_peak_calling.sh    
#SBATCH -N 1
#SBATCH -n 1
#SBATCH -c 2
#SBATCH -p amd
#SBATCH --qos=general
#SBATCH --mem=32G
#SBATCH --mail-type=ALL
#SBATCH --mail-user=ssun@uchc.edu
#SBATCH -o chip_peak_calling.sh_%j.out
#SBATCH -e chip_peak_calling.sh_%j.err

hostname
module load macs3

mkdir temp_macs
directory=/home/FCAM/ssun/GATA3_ChIP_PRO_July2023/ChIP_final/sorted.bam_final/
# GATA_CC against IgG
macs3 callpeak --call-summits -t ${directory}*_GATA_CC*sorted.bam -c ${directory}*IgG*sorted.bam -n GATA_ChIP -g hs -q 0.01 --keep-dup all -f BAMPE --nomodel --tempdir temp_macs

#wc -l GATA_ChIP_summits.bed
#99289 
```

## Removing peaks on contigs and within blacklisted regions


